
---
title: "Basic examples of using BasicP functions"
author: "Alex Sanchez and Mercedes Guerrero"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic examples of using BasicP functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r echo=FALSE, include=FALSE}
require(devtools)
install("/home/mguerrero/Dropbox/Practiques-Grau_Biotec-UVic_Mercedes_Guerrero/BasicP")
require(affy)
require(oligo) #ExonStudy
require("limma")
require(gplots)
```

The goal of this Vignette is to illustrate the use of some functions exported from the package.

## `readOrLoad.rawData'Vignette Info`

This function reads data from cel files or loads previously saved data.


```{r message=FALSE}
readCELS <- TRUE 
my.targets <- "./celfiles/targets.txt"
targets<- read.table("./celfiles/targets.txt", head=TRUE, sep="\t", row.names = 1)
my.fileNames <-paste("./celfiles/",rownames(targets),sep="")
rawDataFileName <- "rawData.Rda"
my.outputDir <- "."
isExonStudy <- FALSE
orgPackage <- "org.Hs.eg" # Necesita este parametro para funciones internas
rawData <- BasicP::readOrLoad.RawData(readCELS = readCELS, phenoDat = my.targets,
                      fileNames = my.fileNames, dataFName =rawDataFileName,
                      outputDir = my.outputDir, exonSt = isExonStudy)
```

## `createOrLoadAnnotations'Vignette Info`

This function creates annotation files or loads previously saved data.

```{r message=FALSE}
loadAnnotations <- FALSE
chipPackAvailable <- TRUE
platformDesignPackAvailable <- FALSE
chipPackage <- "hgu133a2"
platformDesignPackage <- NULL
outputDir <- "./ResultsDir"
annotationsFileName <- "Annotations"
entrezTableFileName <-"Entrezs.Rda"
symbolsTableFileName <-"Symbols.Rda"
controlsTableFileName <- "controls.Rda"

anotacions <- BasicP::createOrLoadAnnotations (loadAnnotations= loadAnnotations, chipPackAvailable = chipPackAvailable, platformDesignPackAvailable = platformDesignPackAvailable,chipPackage = chipPackage, platformDesignPackage = platformDesignPackage, outputDir = outputDir,annotationsFileName = annotationsFileName,entrezTableFileName = entrezTableFileName, symbolsTableFileName = symbolsTableFileName, controlsTableFileName = controlsTableFileName)

```

## `normalization'Vignette Info`

function to normalize the raw data.

```{r message=FALSE , warning=FALSE}
load("rawData.Rda")
rawData <- my.raw
normMethod <- "RMA"
my.targets <- read.AnnotatedDataFrame("./celfiles/targets.txt", header = TRUE, row.names = 1)
celFilesDir <-"./celfiles"
loadFile <- FALSE 
normalized.eset.FileName <-  "normalizedData.Rda"   
outputDir <- "./ResultsDir"
exonStudy <- FALSE

eset_norm <- BasicP::normalization(my.data = rawData, method = normMethod, targetsinfo = my.targets, inputDir = celFilesDir, loadFile = loadFile , normalizedFName = normalized.eset.FileName, outputDir = outputDir, exonSt = exonStudy)
```

## `normplots2File'Vignette Info`

Function that makes a report of the normalization

```{r message=FALSE,warning=FALSE}
load("./ResultsDir/normalizedData.Rda")
repes <- duplicated(exprs(my.norm), MARGIN=1)
exprs(my.norm) <- exprs(my.norm)[!repes,]
eset_norm <- my.norm
my.colors <- rainbow(length(sampleNames(eset_norm)))
my.names <- pData(eset_norm)$ShortName
myCex<- 0.8
dim3 <- FALSE
fileName <- "NormalizedPlots.pdf"
outputDir <- "./ResultsDir"
PCAPlots <- TRUE
csv <- "csv2"

BasicP::normplots2File(my.data = eset_norm, sampleNames = my.names, my.colors = my.colors, my.groups = pData(eset_norm)$Group, my.method = "average",my.cex = myCex ,posText = 2, dim3 = FALSE,fileName = fileName, outputDir = outputDir,PCAPlots = TRUE, csv = fileType)

```

## `filterData'Vignette Info`

Function to filter the data

```{r}
load("./ResultsDir/normalizedData.Rda")
repes <- duplicated(exprs(my.norm), MARGIN=1)
exprs(my.norm) <- exprs(my.norm)[!repes,]
eset_norm <- my.norm
load("./ResultsDir/controls.Rda")
removeNAs <- TRUE
load("./ResultsDir/Entrezs.Rda")
entrezs <- entrezTable
SignalFilter <- TRUE
signalThreshold <- 50
signalFilter.Function <- "filt.by.Signal"
signalThreshold.as.percentage <- TRUE
VarFilter <- TRUE 
variabilityThreshold <- 50
variability.Function <- "sdf"
variabilityThreshold.as.percentage <- TRUE
pairing.Function <- NULL
my.targets <-read.AnnotatedDataFrame("./celfiles/targets.txt", header = TRUE, row.names = 1)
doReport <- TRUE
outputDir <- "./ResultsDir"
FilteringReportFileName <- "FilteringReport.txt"

exprs.filtered <- BasicP::filterData(expres = exprs(eset_norm),controls = names(controlsTable),removeNAs = TRUE, entrezs = entrezs ,bySignal = SignalFilter,signalThr = signalThreshold, grups = pData(eset_norm)$grupo, sigFun.Name = signalFilter.Function, sigThr.as.perc = signalThreshold.as.percentage, byVar = VarFilter, variabilityThr = variabilityThreshold, varFun.Name = variability.Function,
varThr.as.perc = variabilityThreshold.as.percentage, pairingFun.Name = pairing.Function, targets = my.targets, doReport = doReport, outputDir = outputDir, filteringReportFName = FilteringReportFileName)
```

## `saveData'Vignette Info`

Function to save R objects in files.

```{r}
load("./ResultsDir/normalizedData.Rda")
repes <- duplicated(exprs(my.norm), MARGIN=1)
exprs(my.norm) <- exprs(my.norm)[!repes,]
eset_norm <- my.norm
normalized.all.FileName <- "normalized.all"
fileType <-"csv2"
symbolsTable <- load("./ResultsDir/Symbols.Rda")
expres.all.FileName <- "expres.Rda"
linksFileName <- "Links.txt"
outputDir <- "./ResultsDir"

BasicP::saveData(expres = exprs(eset_norm), expres.csv.FileName = normalized.all.FileName, csvType=fileType, description = "Normalized values for all genes", anotPackage = NULL, symbolsVector = symbolsTable, SYMBOL = "SYMBOL", expres.bin.FileName = expres.all.FileName, linksFile = linksFileName, outputDir = outputDir)

BasicP::saveData(expres = exprs.filtered, expres.csv.FileName = "normalized.filtered", csvType = fileType, description = "Normalized values for filtered genes", anotPackage = NULL, symbolsVector = symbolsTable, SYMBOL = "SYMBOL", expres.bin.FileName = "exprs.filtered.Rda",  linksFile = linksFileName, outputDir = outputDir)

```

## `lmAnalysis'Vignette Info`

Setting design and contrasts matrix to proceed with the linear model analysis.

```{r}
targets <- read.table(file ="./celfiles/targets.txt" , header = TRUE, sep = "\t")
column2design<- 5
lev <- targets[,column2design] 
design <- model.matrix( ~ 0 + lev)        
colnames(design) <- levels(lev)
rownames(design) <- targets$ShortName
numParameters <- ncol(design)
print(design)

cont.matrix<-makeContrasts(AvsB= (A -B), AvsL= (A-L), BvsL=(B-L),levels=design)
print(cont.matrix)

```

Linear model analysis.Function included in dolmAnalysis.

```{r}
load("./ResultsDir/exprs.filtered.Rda")
contrasts2test <- 1:ncol(cont.matrix)
anotPackage = NULL
comparison =  "Estudi"
outputDir = "./ResultsDir"
ENTREZIDs = "entrezTable"
SYMBOLIDs = "symbolsTable"
linksFile = "Links.txt"
fitFileName = "fit.Rda"
csvType= "csv"
rows2HTML= NULL
anotFileName <- "Annotations"
runMulticore = 0 
toTIFF= FALSE

fitMain <- BasicP::lmAnalysis(exprs.filtered = exprs.filtered, design = design, cont.matrix = cont.matrix, contrasts2test = contrasts2test, anotPackage = anotPackage, outputDir = outputDir, comparison = comparison, Expressions_And_Top = TRUE , showParams = FALSE , use.dupCorr = FALSE, block = NULL, nDups = 1 , ENTREZIDs = ENTREZIDs, SYMBOLIDs = SYMBOLIDs, linksFile = linksFile,fitFileName = fitFileName , csvType=csvType, rows2HTML = NULL, anotFileName = anotFileName)


```

```{r echo=FALSE, include=FALSE}
#Function to make a list for make the dolmAnalysis.
add2parsList <- function(oneList,object)
{
  pos <- length(oneList) + 1
  oneList[[pos]] <- object
  names(oneList)[pos] <- oneList[[pos]]$comparisonName
  return(oneList)
}

```


## `doLmAnalysis'Vignette Info`
Function to analyze the linerar model with the different parameters.

```{r}
lmParsList <- list()
Estudi <- list(dades = NULL,
               expresFileName = "exprs.filtered.Rda",
               targets = targets,
               designMat = design,
               contMat = cont.matrix,
               whichContrasts = 1:ncol(cont.matrix),
               anotPack = NULL,
               outputDir = outputDir,
               ExpressionsAndTop = TRUE,
               showLmParams = FALSE, 
               use.dupCorr = FALSE,
               block = NULL,
               nDups = 1,
               comparisonName = comparison,  
               ENTREZIDs = "entrezTable",
               SYMBOLIDs = "symbolsTable",
               fileOfLinks = linksFile,
               fitFileName = fitFileName,
               csvType=csvType,
               rows2HTML = NULL,
               anotFilename = anotFileName
               )

lmParsList <- add2parsList(lmParsList, Estudi)
               
for(ix in 1:length(lmParsList))
{
  fit.Main   <- BasicP::doLmAnalysis(lmParsList[ix])
}
```

## `GeneAnnotation'Vignette Info`
Function that annotates genes. Inside doGeneAnnotation


```{r}
genes2annotate <- entrezs[unique(rownames(fitMain$p.value))]

GeneAnnotation(egIDs = genes2annotate, anotPackage = "org.Hs.eg", toHTML = TRUE, outputDir = outputDir, filename = "Annotations",  myTitle = "Annotations for all genes analyzed", specie = "homo sapiens", info2show = c( "Affymetrix", "EntrezGene", "GeneSymbol", "GeneName", "KEGG", "GO"), linksFile = linksFile, maxGenes = NULL)
```

## `doGeneAnnotation'Vignette Info`
Function to annotate the genes.


```{r}
organisme = "hsa"
anotList <- list(fitMain = NULL, fitFileName = fitFileName, my.IDs = entrezs, anotPackage = "org.Hs.eg", toHTML = TRUE, outputDir = outputDir, anotFilename = "Annotations", titleAnotations = "Annotations for all genes analyzed", specie = "homo sapiens", info2show = c( "Affymetrix", "EntrezGene", "GeneSymbol", "GeneName", "KEGG", "GO"), linksFile = linksFile, numGenesPerPage = NULL)

doGeneAnnotation(anotList)
```

## `multipleComp'Vignette Info`
Function to make multiple comparations. Inside doMultCompAnalysis.

```{r}

compName <- c("Group1")
wCont <- 1:3
pValCutOff <- c(0.01)
adjMethod <- c("none")
minLogFoldChange <- c(1)
load("./ResultsDir/Symbols.Rda")
titleText = paste("for", ifelse(adjMethod=="none","p-values","adj. p-values"), "<", pValCutOff, "and |logFC| >", minLogFoldChange, sep = " ")
geneListFName = paste("geneList",compName,ifelse(adjMethod=="none","pvalues","adj-pvalues"),"LT",pValCutOff,"Rda",sep = ".")

geneList <-  BasicP::multipleComp(fitMain = fitMain, whichContrasts = wCont, comparisonName = compName, titleText = titleText, outputDir = outputDir,  anotPackage = "org.Hs.eg", my.symbols = symbolsTable, linksFile = linksFile, multCompMethod = "separate", adjustMethod = adjMethod, selectionType = "any", P.Value.cutoff = pValCutOff, plotVenn = TRUE, colsVenn = NULL, vennColors = c("red","yellow","green","blue","pink") , cexVenn = 1, geneListFName=geneListFName, csvType=csvType, minLFC=minLogFoldChange)

```

## `doMultCompAnalysis'Vignette Info`
Function to perform Multi comparations analysis.


```{r}
mcParsList <- list()
compName <- c("Group1", "Group2")
wCont <- 1:ncol(cont.matrix)
pValCutOff <- c(0.01, 0.01) 
adjMethod <- c("none", "none")
minLogFoldChange <- c(1, 1) 
for(i in 1:length(compName))
{  
mci <- list(fitMain = fitMain,
              fitFileName = fitFileName,
              whichContrasts = wCont[[i]],
              comparisonName = compName[i],
              titleText = paste("for",ifelse(adjMethod[i]=="none",
              "p-values","adj. p-values"), "<", pValCutOff[i], "and |logFC| >",
               minLogFoldChange[i], sep = " "),
             
              anotPackage = "org.Hs.eg",
              my.symbols = symbolsTable,
              outputDir = outputDir,
              fileOfLinks = linksFile,
              multCompMethod = "separate",
              adjustMethod = adjMethod[i], 
              selectionType = "any",
              P.Value.cutoff = pValCutOff[i],
              plotVenn = TRUE,
              colsVenn = NULL,
	      vennColors= c("red","yellow","green","blue","pink"),  
              cexVenn = 1,
              geneListFName = paste("geneList",compName[i],
              ifelse(adjMethod[i]=="none","pvalues","adj-pvalues"),
              "LT",pValCutOff[i],"Rda",sep = "."),
	      
              minLogFC = minLogFoldChange[i],
              csvType = csvType)

  mcParsList <- add2parsList(mcParsList, mci)
}

for(ix in 1:length(mcParsList))
{
  geneList.MCi <- BasicP::doMultCompAnalysis(mcParsList[ix])   
}
```

## `clusterAnalysis'Vignette Info`
Function to draw the clusters.Inside doClusterAnalysis.

```{r}
s2clust <-  which(as.logical(apply(design[,as.logical(apply(abs(as.matrix(cont.matrix[,wCont[[i]]])),1,sum))],1,sum)))
geneListFName = paste("geneList",  compName[i],     ifelse(adjMethod[i]=="none","pvalues","adj-pvalues"), "LT", pValCutOff[i], "Rda", sep = ".")
pal <- colorpanel(n = 32, low = "green", mid = "white", high = "magenta")
load("./ResultsDir/geneList.Group1.pvalues.LT.0.01.Rda")
 clust <- BasicP::clusterAnalysis(expres = exprs.filtered,
                           genes = geneList, 
                           samples = s2clust, 
                           sampleNames =       as.character(targets$ShortName)[s2clust],
                           comparisonName = "Compar 1",
                           anotPackage = "org.Hs.eg",
                           my.symbols = symbolsTable,
                           outputDir = outputDir, 
                           fileOfLinks = linksFile,                              
                           numClusters = 2,
                           rowDistance = NULL,
                           colDistance = NULL,
                           RowVals = TRUE,
                           ColVals = FALSE,
                           escala = "row",
                           colorsSet = pal,
                           densityInfo = "density",
                           colsForGroups = c("pink","pink","pink","pink","pink","blue","blue","blue","blue","blue"),
                           cexForColumns = 0.8,
                           cexForRows = 0.8,
                           Title = paste("Compar 1 with",
                            ifelse(adjMethod[i]=="none","pvalues","adj-pvalues"),
                            "<",
                            pValCutOff[i], ifelse(minLogFoldChange[i]==0, "", paste("\n and |logFC|>=", minLogFoldChange[i], sep=""))),
                           csvType = "csv2")
```

## `doClusterAnalysis'Vignette Info`
Function to draw the clusters.


```{r}
clustParsList <- list()
for(i in 1:length(compName))
{
clustPar <- list(expres = NULL,
                 expresFileName = "exprs.filtered.Rda",
                 geneListFName = paste("geneList",  compName[i],     ifelse(adjMethod[i]=="none","pvalues","adj-pvalues"), "LT", pValCutOff[i], "Rda", sep = "."),
                genes2cluster = NULL, 
                samples2cluster = s2clust, 
                sampleNames = as.character(targets$ShortName)[s2clust], 
                comparisonName = compName[i], 
                anotPackage = "org.Hs.eg",
                my.symbols = symbolsTable,
                outputDir = outputDir,
                fileOfLinks = linksFile,
                numClusters = 2,
                rowDistance = NULL,
                colDistance = NULL,
                RowVals = TRUE,
                ColVals = FALSE,
                escala = "row",
                colorsSet = pal,
                densityInfo = "density",
                colsForGroups = c("pink","pink","pink","pink","pink","blue","blue","blue","blue","blue"),
                cexForColumns = 0.8,
                cexForRows = 0.8,
                Title = paste(compName[i],
                                 "with",
                                 ifelse(adjMethod[i]=="none","pvalues","adj-pvalues"),
                                 "<",
                                 pValCutOff[i], ifelse(minLogFoldChange[i]==0, "", paste("\n and |logFC|>=", minLogFoldChange[i], sep=""))),
                   paste("Comparison:", compName[i], sep=" "),
                   csvType = csvType)
  
clustParsList <- add2parsList(clustParsList, clustPar)
}
  
for(ix in 1:length(clustParsList))
  {
    hm.Estudi <- BasicP::doClusterAnalysis(clustParsList[ix])
  }
  
```

